Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-06 ms.tex
kinetic heating section

\begin{eqnarray}
  \left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}}
  &\approx& 12\eta_{0.02}^{-1} M_{\bullet,8}^{-1.1}(\epsilon_{\rm
    Ia}/0.4) (\tau_{\star}/t_{\rm h})^{-0.56} \nonumber \\
  &\approx& 12 M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4) (\tau_{\star}/t_{\rm h})^{0.74}
\label{eq:rIars}
\end{eqnarray}
If $r_{\rm Ia} \gg r_{\rm s}$, then Ia heating can only be
approximated as a steady heating source near the stagnation radius for
extremely massive SMBHs with $M_{\bullet} \gtrsim 10^9 \Msun$ or in
the case of a very young stellar population ($\tau_{\star} \ll t_{\rm
h}$).  On the other hand, equation (\ref{eq:rIars}) may underestimate
the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be
substantially smaller than its value estimated using $v_{w}^{\rm Ia}$
if the heating rate excluding SN Ia is less than this.

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between subsequent supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

>>>>

\begin{eqnarray}
  \left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}}
  &\approx& 12\eta_{0.02}^{-1} M_{\bullet,8}^{-1.1}(\epsilon_{\rm
    Ia}/0.4) (\tau_{\star}/t_{\rm h})^{-0.56} \nonumber \\
  &\approx& 12 M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4) (\tau_{\star}/t_{\rm h})^{0.74}
\label{eq:rIars}
\end{eqnarray}
If $r_{\rm Ia} \gg r_{\rm s}$, then Ia heating can only be
approximated as a steady heating source near the stagnation radius for
extremely massive SMBHs with $M_{\bullet} \gtrsim 10^9 \Msun$ or in
the case of a very young stellar population ($\tau_{\star} \ll t_{\rm
h}$).

%On the other hand, equation (\ref{eq:rIars}) may underestimate
%the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be
%substantially smaller than its value estimated using $v_{w}^{\rm Ia}$
%if the heating rate excluding SN Ia is less than this.

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

As in the case of Compton heating, kinetic heating could in principle  `self-regulate' the accretion flow insofar as a lower heating rate results in a higher accretion rate (eq.~[\ref{eq:vjet}]), which in turn may create stronger jet feedback.  However, as it is not possible to make concrete predictions for such a process given the uncertainty in the efficiency of jet heating, we defer a more detailed discussion of the possible role of kinetic feedback to $\S\ref{sec:heating}$.
